Confined Cubic Blue Phases under Shear

### 2011-11-11 papers/lmcproc/lmcproc.tex
Various minor changes to text.
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1486 40876441-3496-0410-a613-a2fca35b2ae2

\address{$^1$ Centre for Computational Science, University College London, UK\\$^2$ Edinburgh Parallel Computing Centre, University of Edinburgh, UK\\$^3$ School of Physics and Astronomy, University of Edinburgh, UK}
\ead{o.henrich@ucl.ac.uk}

\begin{abstract}
We study the behaviour of confined cubic blue phases under a shear flow, via lattice Boltzmann simulations. We focus on the two experimentally observed
phases, blue phase I and blue phase II. The disclination network of blue phase II continuously breaks and reforms under shear, 
leading to an oscillatory stress response in time. The oscillations are only regular for very thin samples. For thicker samples, the shear leads to a ``stick-slip'' motion
of part of the network along the vorticity direction. Blue phase I responds vastly differently: its defect network undergoes seemingly
chaotic rearrangements under shear, irrespective of system size. 
\end{abstract}

@@

\section{Introduction}
% Put \label in argument of \section for cross-referencing
%\section{\label{}}
Cholesterics are liquid crystals in which the local director field spontaneously twist in thermodynamic equilibrium~\cite{deGennes}.
The preferred configuration close to the isotropic boundary features twist around two perpendicular axes, as opposed to just one axis in the regular cholesteric state, and the corresponding deformation is denoted a ``double-twist cylinder''.
As it is topologically impossible to continuously patch 3D space with double-twist cylinders, defects arise. These may organise into a variety of regular periodic lattices, giving rise to the so-called cubic blue phases (BPs) \cite{Grebel:1984,Wright:1989}. There are two experimentally observed cubic blue phases, BPI and BPII (a third, BPIII, is likely amorphous~\cite{Henrich:2011a}).\\
BPs have been long considered as purely of academic interest due to their very narrow range of stability, but this has drastically changed since the creation of polymer- and temperature-stabilised BPs~\cite{Kikuchi:2002,Coles:2005}, which has opened up the possibility of novel applications.
During the last few years considerable progress has been achieved regarding the behaviour of BPs in confined geometries~\cite{Fukuda:2010a, Fukuda:2010b, Ravnik:2011b}, under external fields \cite{Alexander:2008,Fukuda:2009,Henrich:2010a,Castles:2010,Tiribocchi:2011}, and in the presence of colloidal particles \cite{Ravnik:2011a}. 
The kinetics of BP domain growth have been recently addressed in~\cite{Henrich:2010b}. However, our understanding of their dynamical behaviour under flow remains
very limited. The aim of this work is to contribute to fill this gap by studying, for the first time, the response of confined BP samples to a shear flow. \\
Flow in cholesterics is both strongly non-Newtonian and highly anisotropic. If a cholesteric helix is subjected to, for instance, a Poiseuille flow along the helical axis, for small pressure differences flow occurs mainly through ``permeation'', first investigated by Helfrich \cite{Helfrich:1969}. In the permeation mode the liquid crystal flows by leaving the director field virtually unchanged, and this leads to high dissipation and large viscosities. Marenduzzo et al. \cite{Marenduzzo:2006a,Marenduzzo:2006b} simulated shear and Poiseuille flow in cholesteric liquid crystals in the permeation mode, and showed the importance of the boundary conditions in determining the apparent viscosity of the fluid. They also found that a strong secondary flow appears.
Rey \cite{Rey:1996a, Rey:1996b} studied shear in cholesterics oriented with the helix along the vorticity axis and found that at low Ericksen number travelling twist waves appear which lead to the rotation of the cholesteric helix. At higher forcing, the helix uncoils and leaves a flow-induced nematic phase. 
Rey also studied cholesterics subjected to steady and low frequency small amplitude oscillatory shear for different helix orientations \cite{Rey:2000, Rey:2002}, where he found that splay–bend–twist deformations were excited when the helix is aligned along the flow direction. Splay–bend deformation occured when the helix is aligned along the velocity gradient, and only twist deformations when the helix is aligned along the vorticity axis. \\
Dupuis et al. \cite{Dupuis:2005} performed the first numerical investigation of BP rheology in Poiseuille flow, starting from equilibrium structures of BPI and BPII and a periodic array of doubly twisted cylinders.
Under small forcing, the network opposed the flow giving rise to a significant increase in apparent viscosity.
Upon increasing the forcing they found clear evidence of shear thinning.
In the crossover region they predicted a novel oscillatory regime where the network continuously breaks and reforms as portions of the disclinations in the center of the channel move to neighboring cells and relink with the parts of the network left behind by the flow. With respect to the cholesteric case, the viscosity still decreases with forcing (the system shear thins) but much less than in cholesterics in the permeation mode, which is in agreement with experiments experiments \cite{Zapotocky:1999, Ramos:2002}.\\
This paper is organised as follows: 
In Sec. \ref{sec_model} we introduce the Beris-Edwards model for hydrodynamics of liquid crystals, which is a generalisation of the Eriksen and Leslie's theory of nematodynamics~\cite{deGennes}.
In Sec. \ref{sec_results} we describe the simulation methodology and boundary conditions we applied.
Results of a confined BPII in rectilinear shear flow are presented in Sec. \ref{sec_bpii} with particular focus on the dependence of the thickness of the sample and thermodynamic state.
In Sec. \ref{sec_bpi} we show corresponding results for BPI, which indicates fundamental differences in the flow behaviour of BPI and BPII.

@@

Our approach is based on a well-established model for the hydrodynamics of cholesteric liquid crystals \cite{Beris:1994,Olmsted:1999} that describes their state in terms of a traceless, symmetric tensor order parameter ${\mathbf Q}({\mathbf r})$. 
In the uniaxial approximation the order parameter is given by $Q_{\alpha \beta}=\langle n_\alpha n_\beta - 1/3\; \delta_{\alpha\beta}\rangle$ with ${\mathbf n}$ the director field.
Brackets indicate mesoscopic averaging over a large number of individual molecules.
The largest eigenvalue of ${\mathbf Q}$, $0\le q_s\le\frac{2}{3}$ characterizes the local degree of orientational order.
The thermodynamical properties of the liquid crystal are determined by a free energy
${\cal F}$, whose density $f$ consists of a bulk contribution $f_b$ and a gradient part $f_g$, as follows,
\begin{eqnarray}
f_b&=&\frac{A_0}{2}\left(1-\frac{\gamma}{3}\right) Q_{\alpha \beta}^2-\frac{A_0 \gamma}{3}Q_{\alpha \beta} Q_{\beta \gamma} Q_{\gamma \alpha}+\frac{A_0 \gamma}{4}(Q_{\alpha \beta}^2)^2\nonumber\\
f_g&=&\frac{K}{2}(\varepsilon_{\alpha\gamma\delta} \partial_\gamma Q_{\delta\beta}+2 q_0 Q_{\alpha \beta})^2+\frac{K}{2}(\partial_\beta Q_{\alpha \beta})^2\label{FE}
\end{eqnarray}
The first term contains the bulk-free energy constant $A_0$ and the inverse temperature $\gamma$ which controls the magnitude of order.
The second part quantifies the cost of elastic distortions, which are proportional to the elastic constant $K$ --
we work for simplicity in the one-elastic constant approximation. The wavevector $q_0=2\pi/p_0$, where $p_0$ is the cholesteric pitch.
The actual periodicity of the BP structure, $p$, does not need to be equal to $p_0$.
The ``redshift'' $r=p/p_0$ is adjusted during the simulation by following a simple procedure similar to \cite{Alexander:2006}.\\
A particular thermodynamic state is specified by two dimensionless quantities: the effective temperature $\tau$ and chirality $\kappa$,
which are given by
\begin{eqnarray}
\tau&=&\frac{27(1-\gamma/3)}{\gamma}\nonumber\\
\kappa&=&\sqrt{\frac{108 K q_0^2}{A_0 \gamma}}\nonumber.
\end{eqnarray}

The dynamical evolution of the order parameter is given by the equation 
\begin{equation}
\left(\partial_t+ u_\alpha \partial_\alpha \right){\mathbf Q} - {\mathbf S}({\mathbf W},{\mathbf Q}) = \Gamma {\mathbf H}.
\label{eqn2}
\end{equation}
The first term on the left hand side of Eq.\ref{eqn2} is a material derivative, which describes the rate of change of a quantity moving along with the flow.
The second term accounts for the rate of change due to local velocity gradients $W_{\alpha \beta}=\partial_\beta u_\alpha$,
\begin{eqnarray}
{\mathbf S}({\mathbf W}, {\mathbf Q}) &=& (\xi {\mathbf A} + {\boldsymbol \Omega})({\mathbf Q}+\frac{\mathbf I}{3})\nonumber\\
& &\hspace*{-1.5cm}+ ({\mathbf Q}+\frac{\mathbf I}{3})(\xi {\mathbf A}  - {\boldsymbol \Omega})-2 \xi ({\mathbf Q}+\frac{\mathbf I}{3})Tr({\mathbf Q W}),
\label{eqn3}
\end{eqnarray}
where $Tr$ denotes the tensorial trace and ${\mathbf A}=({\mathbf W}+{\mathbf W}^T)/2$ and ${\boldsymbol \Omega}=({\mathbf W}-{\mathbf W}^T)/2$ are the symmetric and antisymmetric part of the velocity gradients, respectively. $\xi$ is a constant depending on the molecular details of the liquid crystals.
Flow alignment occurs if $\xi \cos{2\theta}=\frac{3q_s}{2+q_s}$ has a real solution, where $\theta$ is the Leslie-angle: we select this case by setting $\xi=0.7$ in our runs.
${\mathbf H}$ is the molecular field, which is a functional derivative of $\cal F$ that preserves the tracelessness of $\mathbf Q$.
\begin{equation}
{\bf H}=-\frac{\delta {\cal F}}{\delta {\bf Q}}+\frac{\bf I}{3}\, Tr\left(\frac{\delta {\cal F}}{\delta {\bf Q}}\right).
\label{eqn4}
\end{equation}
The rotational diffusion constant $\Gamma$ is proportional to the inverse of the rotational viscosity $\gamma_1=2 q_s^2/\Gamma$.\\
The time evolution of density and velocity are respectively governed by the continuity and the following 
Navier-Stokes equation Eq.\ref{eqn6}, 
\begin{eqnarray}
\partial_t u_\alpha +\rho \,u_\beta \partial_\beta u_\alpha&=&\partial_\beta \Pi_{\alpha \beta}+\eta\, \partial_\beta \{ \partial_\alpha u_\beta + \partial_\beta u_\alpha +(1+3\frac{P_0}{\partial \rho} )\partial_\mu u_\mu \delta_{\alpha \beta})\}. 
\label{eqn6}
\end{eqnarray}

At low flow rates the fluid can be considered as incompressible, so that the last term on the righthand side of Eq.\ref{eqn6} can be safely neglected.
$\eta$ is an isotropic background viscosity which is set to $\eta=0.8333$ in our case.
The pressure tensor reads explicitly
\begin{eqnarray}
\Pi_{\alpha \beta}&=&P_0-\xi H_{\alpha \gamma}\left(Q_{\gamma \beta} +\frac{1}{3} \delta_{\gamma \beta}\right)-\xi \left(Q_{\alpha \gamma} +\frac{1}{3} \delta_{\alpha \gamma}\right) H_{\gamma \beta}\nonumber\\
&+& 2 \xi  \left(Q_{\alpha \beta} +\frac{1}{3} \delta_{\alpha \beta}\right) Q_{\gamma \nu} H_{\gamma \nu}-\partial_\alpha Q_{\gamma \nu} \frac{\delta{\cal F}}{\delta \partial_{\beta} Q_{\gamma \nu}}\nonumber\\
&+&Q_{\alpha \gamma}H_{\gamma \beta}-H_{\alpha \gamma} Q_{\gamma \beta}.
\label{eqn7}
\end{eqnarray}
In the isotropic state when ${\bf Q}\equiv 0$ Eq.\ref{eqn7} is reduced to the scalar pressure, which is constant to a very good approximation.\\ 
The system of coupled partial differential equations is solved by means of a hybrid scheme \cite{Marenduzzo:2007} that uses a lattice Boltzmann algorithm with predictor-corrector scheme for the continuity equation and Eq.\ref{eqn6} and a finite difference scheme for the equation of motion of the tensor order parameter Eq.\ref{eqn2}.
More details on the algorithm can be found in \cite{Denniston:2001, Denniston:2004}.

@@

Shear flow was then started by applying shear boundary conditions to the LB distribution functions as in~\cite{Denniston:2004}. As regards ${\mathbf Q}$, the equilibrated order parameter was advected on the top and bottom boundary, according to the imposed wall velocities. Periodic boundary conditions were retained along the flow direction and the direction of vorticity.
This set-up is similar to the one Dupuis et al. used to simulate Poiseuille flow \cite{Dupuis:2005}. The pinning of the defect network at the boundaries may be realised in practice by impurities or surface defects.
During shear flow the unit cell size was kept fixed and a constant redshift equal to the value at the end of the equilibration protocol was assumed.\\ 
The total number of unit cells in our simulations was $n_x\times n_y \times n_z= 1, \dots, 4 \times 4 \times 4$ for both blue phases, using a resolution of 32 lattice Boltzmann units (LBU) per pitch length (16 per unit cell) for BPII and 64 LBU per pitch length (32 per unit cell) for BPI.  
This provided enough spatial resolution to track even complex rearrangements of the disclination networks.
A timestep and lattice spacing in LBU corresponds to $\sim 1 {\rm ns}$ and $\sim 10{\rm nm}$ in SI-units.
The LBU of stress equals to about $10^8 Pa$. 
More details about the conversion from LBU to SI-units can be found in \cite{Henrich:2010b,Henrich:2011a}.

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.24\textwidth]{bp2_t-0o5_k2_25k_y.png}
\includegraphics[width=0.24\textwidth]{bp2_t-0o5_k2_50k_y.png}
\includegraphics[width=0.24\textwidth]{bp2_t-0o5_k2_75k_y.png}
\includegraphics[width=0.24\textwidth]{bp2_t-0o5_k2_100k_y.png}
\caption{Shear flow of BPII at different gap sizes: The pictures show a section of the entire disclination network of $n_x=1 \mbox{ (yellow)}, 2 \mbox{ (green)}, 3 \mbox{ (red) and } 4 \mbox{ (blue)}$ unit cells along the gap and $n_y \times n_z =2 \times 2$ unit cells in flow and vorticity direction. The thermodynamic parameters temperature and chirality are $\tau=-0.5$ and chirality $\kappa=2$, respectively, where BPII is the equlibrium phase. The velocity gradient is oriented along the vertical direction with the top (bottom) wall moving into (out of) the viewing plane. The strain rate is $\dot{\gamma}=7.8125\ex{-5}$ for all gap sizes.}
\label{fig1}
\end{figure}

We first report the results for BPII.
Fig.\ref{fig1} shows the disclination network in steady shear flow for different number of unit cells between the walls.
(We prefer the perspective along the flow direction because the network topology is clearer then.)
Temperature and chirality are $\tau=-0.5$ and $\kappa=2$ respectively, and for these parameters BPII is the equilibrium phase.
The velocity of the walls has been adapted so that the same shear rate $\dot{\gamma}=7.8125\ex{-5}$ is achieved for all gap sizes.
Shear rates four times larger and smaller were also applied and produced very similar results, apart from the flow velocities which scale up accordingly.
For all gap sizes we observe an affine transformation in gradient-flow plane that leads to a break-up of the network at strains of about half a unit cell size.
Shorlty after the break-up the network reforms further downstream, resulting in a periodic, regular structure that is very close to the equilibrium
BPII disclination network. Similar oscillations have been previously reported for
BPs in Poiseuille flow, with a gap of one unit cell size \cite{Dupuis:2005}.\\
It is interesting to compare the results of different gap sizes.
For $n_x=1$ we observe that the BPII-network is slightly deformed with respect to the quiescent state, but reforms always at the same position.
At gap sizes of more than one unit cell the BPII network distorts more under the same shear rate, 
suggesting that the structure within the sample is highly affected by the anchoring at the boundaries.
Interestingly, when the sample is thicker than one unit cell a movement of the network in negative vorticity direction arises, 
which is not observed for $n_x=1$.
This network motion does not proceeds at a constant speed, rather it proceeds in a ``stick-slip'' fashion: elastic stresses first slowly build up, and are then 
quickly released as the network ``slips'' along the vorticity direction.
Changing from a left- to a right-handed helix inverts the direction of motion of the network.
Hence the ``stick-slip'' motion is related to the chiral nature of the helix. This phenomenon may also bear some similarities to
the travelling wave motion in cholesterics sheared along the vorticity direction 
below a critical Ericksen number \cite{Rey:1996a, Rey:1996b}.
It is interesting to note that the secondary flow in the BP
does {\em not} match the direction of the ``stick-slip'' motion of the network, which is then not connected to mass flow
(in this respect this phenomenon is close to a permeation flow).

\begin{figure*}[t]
\centering
\includegraphics[width=0.3\textwidth]{disc_bp2_tk_scan_25k.png}
\includegraphics[width=0.3\textwidth]{disc_bp2_tk_scan_75k.png}
\includegraphics[width=0.3\textwidth]{disc_bp2_tk_scan_125k.png}
\caption{BPII in shear flow for different thermodynamic parameters: The images show the network as it appears along the flow direction at the reference state (blue, $\tau=-0.5, \kappa=2$), a high temperature state (red, $\tau=0.5, \kappa=2$) and a low chirality state (green, $\tau=-0.5, \kappa=1$), where BPII is metastable. The strain rate is $\dot{\gamma}=7.8125\ex{-5}$ in all three cases. The pictures show a central section consisting of 8 unit cells from a larger run with $n_x=4$ gap size.}
\label{fig2}
\end{figure*}

The speed of the network in vorticity direction depends on the thermodynamic state of the blue phase.
A comparison between the state from Fig. \ref{fig1} with a ``high temperature'' state at $\tau=0.5, \kappa=2$ and a ``low chirality state'' at $\tau=-0.5, \kappa=1$
is shown in Fig.\ref{fig2}. Note that in both these states BPII is now a metastable phase. 
The images in Fig.~\ref{fig2} give a sequence of initially overlapping disclination networks that separate during the course of the simulation.
Of all three states the high temperature state is the fastest one, followed by the reference state and the low chirality state.

>>>>

\address{$^1$ Centre for Computational Science, University College London, UK\\$^2$ Edinburgh Parallel Computing Centre, The University of Edinburgh, UK\\$^3$ School of Physics and Astronomy, The University of Edinburgh, UK}
\ead{o.henrich@ucl.ac.uk}

\begin{abstract}
We study the behaviour of confined cubic blue phases under shear flow
via lattice Boltzmann simulations. We focus on the two experimentally observed
phases, blue phase I and blue phase II. The disclination network of blue phase II continuously breaks and reforms under shear, 
leading to an oscillatory stress response in time. The oscillations are only regular for very thin samples. For thicker samples, the shear leads to a ``stick-slip'' motion
of part of the network along the vorticity direction.
Blue phase I responds very differently: its defect network undergoes seemingly
chaotic rearrangements under shear, irrespective of system size. 
\end{abstract}

@@

\section{Introduction}
% Put \label in argument of \section for cross-referencing
%\section{\label{}}
Cholesterics are liquid crystals in which the local director field spontaneously twist in thermodynamic equilibrium~\cite{deGennes}.
The preferred configuration close to the isotropic boundary features twist around two perpendicular axes, as opposed to just one axis in the regular cholesteric state, and the corresponding deformation is denoted a ``double-twist cylinder''.
As it is topologically impossible to patch continuously 3D space with double-twist cylinders, defects arise. These organise into a variety of regular periodic lattices, giving rise to the so-called cubic blue phases (BPs) \cite{Grebel:1984,Wright:1989}. There are two experimentally observed cubic blue phases, BPI and BPII (a third, BPIII, is thought to be amorphous~\cite{Henrich:2011a}).

BPs have been long considered as purely of academic interest due to their very narrow range of stability. This view has changed since the creation of
polymer-stabilised and temperature-stabilised BPs~\cite{Kikuchi:2002,Coles:2005}, which has opened up the possibility of novel applications.
During the last few years considerable progress has been achieved regarding the behaviour of BPs in confined geometries~\cite{Fukuda:2010a, Fukuda:2010b, Ravnik:2011b}, under external fields \cite{Alexander:2008,Fukuda:2009,Henrich:2010a,Castles:2010,Tiribocchi:2011}, and in the presence of colloidal particles \cite{Ravnik:2011a}. 
The kinetics of BP domain growth have been recently addressed in~\cite{Henrich:2010b}. However, our understanding of their dynamical behaviour under flow remains
very limited. The aim of this work is to address this issue by studying,
for the first time, the response of confined BP samples to a shear flow.

Flow in cholesterics is both strongly non-Newtonian and highly anisotropic.
For example, if a cholesteric helix is subjected to a Poiseuille flow along
the helical axis, small pressure differences drive flow mainly through
``permeation'', first investigated by Helfrich \cite{Helfrich:1969}.
In the permeation mode the liquid crystal flows while leaving the director
field virtually unchanged, which leads to high dissipation and large
viscosities. Marenduzzo et al. \cite{Marenduzzo:2006a,Marenduzzo:2006b} simulated shear and Poiseuille flow in cholesteric liquid crystals in the permeation mode, and showed the importance of the boundary conditions in determining the apparent viscosity of the fluid. They also found that a strong secondary flow appears.
Rey \cite{Rey:1996a, Rey:1996b} studied shear in cholesterics oriented with the helix along the vorticity axis and found that, at low Ericksen number, travelling twist waves appear which lead to the rotation of the cholesteric helix. At higher forcing, the helix uncoils and leaves a flow-induced nematic phase. 
Rey also studied cholesterics subjected to both steady flow and low frequency
small amplitude oscillatory shear for different helix orientations
\cite{Rey:2000, Rey:2002}. He found that splay/bend/twist deformations were
excited when the helix was aligned along the flow direction, splay/bend
deformation occured when the helix was aligned along the velocity gradient,
but only twist deformations
appeared when the helix wass aligned along the vorticity axis.

Dupuis et al. \cite{Dupuis:2005} performed the first numerical investigation of BP rheology in Poiseuille flow, starting from equilibrium structures of BPI and BPII and a periodic array of doubly twisted cylinders.
Under small forcing, the network opposed the flow giving rise to a significant increase in apparent viscosity.
Upon increasing the forcing they found clear evidence of shear thinning.
In the crossover region they predicted a novel oscillatory regime where the network continuously breaks and reforms as portions of the disclinations in the center of the channel move to neighboring cells and relink with the parts of the network left behind by the flow. Compared with the cholesteric case, the viscosity still decreases with forcing (the system shear thins) but much less than in cholesterics in the permeation mode, which is in agreement with experiments
\cite{Zapotocky:1999, Ramos:2002}.

This paper is organised as follows.
In Sec.~\ref{sec_model} we introduce the Beris-Edwards model for hydrodynamics of liquid crystals, which is a generalisation of the Eriksen and Leslie's theory of nematodynamics~\cite{deGennes}.
In Sec.~\ref{sec_results} we describe the simulation methodology and the
boundary conditions we applied.
Results for a confined BPII in rectilinear shear flow are presented in
Sec.~\ref{sec_bpii} with particular focus on the dependence of the flow
on the thickness of the sample and thermodynamic state.
In Sec.~\ref{sec_bpi} we show corresponding results for BPI, which indicates fundamental differences in the flow behaviour of BPI and BPII.

@@

Our approach is based on a well-established model for the hydrodynamics of
cholesteric liquid crystals \cite{Beris:1994,Olmsted:1999}. The model
describes their state in terms of a traceless, symmetric tensor order parameter ${\mathbf Q}({\mathbf r})$. 
In the uniaxial approximation, the order parameter is given by $Q_{\alpha \beta}=\langle n_\alpha n_\beta - 1/3\; \delta_{\alpha\beta}\rangle$ with ${\mathbf n}$ the director field.
Brackets indicate mesoscopic averaging over a large number of individual molecules.
The largest eigenvalue of ${\mathbf Q}$, $0\le q_s\le\frac{2}{3}$ characterizes the local degree of orientational order.
The thermodynamic properties of the liquid crystal are determined by a free energy
${\cal F}$, whose density $f$ consists of a bulk contribution $f_b$ and a gradient part $f_g$, as follows,
\begin{eqnarray}
f_b&=&\frac{A_0}{2}\left(1-\frac{\gamma}{3}\right) Q_{\alpha \beta}^2-\frac{A_0 \gamma}{3}Q_{\alpha \beta} Q_{\beta \gamma} Q_{\gamma \alpha}+\frac{A_0 \gamma}{4}(Q_{\alpha \beta}^2)^2,\nonumber\\
f_g&=&\frac{K}{2}(\varepsilon_{\alpha\gamma\delta} \partial_\gamma Q_{\delta\beta}+2 q_0 Q_{\alpha \beta})^2+\frac{K}{2}(\partial_\beta Q_{\alpha \beta})^2.\label{FE}
\end{eqnarray}
The first term contains the bulk-free energy constant $A_0$ and the inverse temperature $\gamma$ which controls the magnitude of order.
The second part quantifies the cost of elastic distortions, which are proportional to the elastic constant $K$ --
we work for simplicity in the one-elastic constant approximation. The wavevector $q_0=2\pi/p_0$, where $p_0$ is the cholesteric pitch.
The actual periodicity of the BP structure, $p$, does not need to be equal to $p_0$.
The ``redshift'' $r=p/p_0$ is adjusted during the simulation by following a simple procedure similar to \cite{Alexander:2006}.

A particular thermodynamic state is specified by two dimensionless quantities: the effective temperature $\tau$ and chirality $\kappa$,
which are given by
\begin{eqnarray}
\tau&=&\frac{27(1-\gamma/3)}{\gamma}\nonumber\\
\kappa&=&\sqrt{\frac{108 K q_0^2}{A_0 \gamma}}\nonumber.
\end{eqnarray}
The dynamical evolution of the order parameter is given by the equation 
\begin{equation}
\left(\partial_t+ u_\alpha \partial_\alpha \right){\mathbf Q} - {\mathbf S}({\mathbf W},{\mathbf Q}) = \Gamma {\mathbf H}.
\label{eqn2}
\end{equation}
The first term on the left hand side of Eq.\ref{eqn2} is a material derivative, which describes the rate of change of a quantity moving along with the flow.
The second term accounts for the rate of change due to local velocity gradients $W_{\alpha \beta}=\partial_\beta u_\alpha$,
\begin{eqnarray}
{\mathbf S}({\mathbf W}, {\mathbf Q}) &=& (\xi {\mathbf A} + {\boldsymbol \Omega})({\mathbf Q}+\frac{\mathbf I}{3})\nonumber\\
& &\hspace*{-1.5cm}+ ({\mathbf Q}+\frac{\mathbf I}{3})(\xi {\mathbf A}  - {\boldsymbol \Omega})-2 \xi ({\mathbf Q}+\frac{\mathbf I}{3})Tr({\mathbf Q W}),
\label{eqn3}
\end{eqnarray}
where $Tr$ denotes the tensorial trace and ${\mathbf A}=({\mathbf W}+{\mathbf W}^T)/2$ and ${\boldsymbol \Omega}=({\mathbf W}-{\mathbf W}^T)/2$ are the symmetric and antisymmetric part of the velocity gradients, respectively. $\xi$ is a constant depending on the molecular details of the liquid crystal.
Flow alignment occurs if $\xi \cos{2\theta}=\frac{3q_s}{2+q_s}$ has a real solution, where $\theta$ is the Leslie-angle: we select this case by setting $\xi=0.7$ in our simulations.
${\mathbf H}$ is the molecular field, which is a functional derivative of $\cal F$ that preserves the tracelessness of $\mathbf Q$.
\begin{equation}
{\bf H}=-\frac{\delta {\cal F}}{\delta {\bf Q}}+\frac{\bf I}{3}\, Tr\left(\frac{\delta {\cal F}}{\delta {\bf Q}}\right).
\label{eqn4}
\end{equation}
The rotational diffusion constant $\Gamma$ is proportional to the inverse of the rotational viscosity $\gamma_1=2 q_s^2/\Gamma$.\\
The time evolution of density and velocity are respectively governed by the continuity and the following 
Navier-Stokes equation Eq.\ref{eqn6}, 
\begin{eqnarray}
\partial_t u_\alpha +\rho \,u_\beta \partial_\beta u_\alpha
= \partial_\beta \Pi_{\alpha \beta}
+\eta\, \partial_\beta \{ \partial_\alpha u_\beta + \partial_\beta u_\alpha
+(1+3\frac{P_0}{\rho} )\partial_\mu u_\mu \delta_{\alpha \beta}\}. 
\label{eqn6}
\end{eqnarray}

At low flow rates the fluid can be considered as incompressible, so that the last term on the righthand side of Eq.\ref{eqn6} can be safely neglected.
$\eta$ is an isotropic background viscosity which is set to $\eta=0.8333$ in our case.
The pressure tensor reads explicitly
\begin{eqnarray}
\Pi_{\alpha \beta}&=&P_0-\xi H_{\alpha \gamma}\left(Q_{\gamma \beta} +\frac{1}{3} \delta_{\gamma \beta}\right)-\xi \left(Q_{\alpha \gamma} +\frac{1}{3} \delta_{\alpha \gamma}\right) H_{\gamma \beta}\nonumber\\
&+& 2 \xi  \left(Q_{\alpha \beta} +\frac{1}{3} \delta_{\alpha \beta}\right) Q_{\gamma \nu} H_{\gamma \nu}-\partial_\alpha Q_{\gamma \nu} \frac{\delta{\cal F}}{\delta \partial_{\beta} Q_{\gamma \nu}}\nonumber\\
&+&Q_{\alpha \gamma}H_{\gamma \beta}-H_{\alpha \gamma} Q_{\gamma \beta}.
\label{eqn7}
\end{eqnarray}
In the isotropic state ${\bf Q}\equiv 0$ and Eq.\ref{eqn7} is reduced to the scalar pressure, which is constant to a very good approximation.\\ 
The system of coupled partial differential equations is solved by means of a hybrid scheme \cite{Marenduzzo:2007} that uses a lattice Boltzmann algorithm with predictor-corrector scheme for the continuity equation and Eq.\ref{eqn6} and a finite difference scheme for the equation of motion of the tensor order parameter Eq.\ref{eqn2}.
More details on the algorithm can be found in \cite{Denniston:2001, Denniston:2004}.

@@

Shear flow was then started by applying shear boundary conditions to the LB distribution functions as in~\cite{Denniston:2004}. For ${\mathbf Q}$, the equilibrated order parameter was advected on the top and bottom boundary, according to the imposed wall velocities. Periodic boundary conditions were retained along the flow direction and the direction of vorticity.
This set-up is similar to the one Dupuis et al. used to simulate Poiseuille flow \cite{Dupuis:2005}. The pinning of the defect network at the boundaries may be realised in practice by impurities or surface defects.
During shear flow the unit cell size was kept fixed and a constant redshift equal to the value at the end of the equilibration protocol was assumed.

The total number of unit cells in our simulations of both BPI and BPII
was $n_x\times n_y \times n_z= \{1,2,3,4\} \times 4\times 4$.
A resolution of 32 lattice Boltzmann units (LBU) per pitch length
(16 per unit cell) was used for BPII, and 64 LBU per pitch length
(32 per unit cell) used for BPI.  
This provided enough spatial resolution to track even complex rearrangements
of the disclination networks in each case.
A timestep and lattice spacing in LBU corresponds to $\sim 1 {\rm ns}$ and $\sim 10{\rm nm}$ in SI-units.
The LBU of stress equals to about $10^8$~Pa. 
More details about the conversion from LBU to SI-units can be found in \cite{Henrich:2010b,Henrich:2011a}.

@@

\begin{figure}[t]
\includegraphics[width=0.24\textwidth]{bp2_t-0o5_k2_25k_y.png}
\includegraphics[width=0.24\textwidth]{bp2_t-0o5_k2_50k_y.png}
\includegraphics[width=0.24\textwidth]{bp2_t-0o5_k2_75k_y.png}
\includegraphics[width=0.24\textwidth]{bp2_t-0o5_k2_100k_y.png}
\caption{Shear flow of BPII at different gap sizes. The pictures show
snapshots of a subset of the entire disclination network in a system
with walls at top and bottom, and the flow direction into/out of the
page. The different number of unit cells in the velocity gradient
direction have been superposed and are shown in different colours:
$n_x=1 \mbox{ (yellow)}, n_x=2 \mbox{ (green)}, n_x=3 \mbox{ (red) and }
n_x=4 \mbox{ (blue)}$. All cases show $n_y \times n_z =2 \times 2$ unit
cells in flow and vorticity directions. The temperature and chirality are
$\tau=-0.5$ and $\kappa=2$, respectively, where BPII is the equlibrium phase.
The strain rate is $\dot{\gamma}=7.8125\ex{-5}$ in all cases.}
\label{fig1}
\end{figure}

We first report the results for BPII.
Fig.~\ref{fig1} shows the disclination network in steady shear flow for different number of unit cells between the walls.
(We prefer the perspective along the flow direction because the network topology is clearer.)
Temperature and chirality are $\tau=-0.5$ and $\kappa=2$ respectively, and for these parameters BPII is the equilibrium phase.
The velocity of the walls has been adapted so that the same shear rate $\dot{\gamma}=7.8125\ex{-5}$ is achieved for all gap sizes.
Shear rates four times larger and smaller were also applied and produced very similar results, apart from the flow velocities which scale up accordingly.
For all gap sizes we observe an affine transformation in gradient-flow plane that leads to a break-up of the network at strains of about half a unit cell size.
Shorlty after the break-up the network reforms further downstream, resulting in a periodic, regular structure that is very close to the equilibrium
BPII disclination network. Similar oscillations have been previously reported for
BPs in Poiseuille flow, with a gap of one unit cell size \cite{Dupuis:2005}.

It is interesting to compare the results for different gap sizes.
For $n_x=1$ we observe that the BPII-network is slightly deformed with
respect to the quiescent state, but always reforms at the same position.
For $n_x > 1$, the BPII network distorts more under the same shear rate, 
suggesting that the structure within the sample is strongly affected by
the anchoring at the boundaries.
Interestingly, when the sample is thicker than one unit cell, a movement
of the network in the negative vorticity direction arises
which is not observed for $n_x=1$.
This network motion does not proceed at a constant speed, but rather
proceeds in a ``stick-slip'' fashion: elastic stresses first slowly build up,
and are then quickly released as the network ``slips'' along the vorticity
direction. Changing from a left-handed to a right-handed helix inverts the
direction of motion of the network.
Hence the ``stick-slip'' motion is related to the chiral nature of the helix.
This phenomenon may bear some similarity to the travelling wave motion in
cholesterics sheared along the vorticity direction 
below a critical Ericksen number \cite{Rey:1996a, Rey:1996b}.
It is interesting to note that the secondary flow in the BP
does {\em not} match the direction of the ``stick-slip'' motion of the network, which is then not connected to mass flow
(in this respect this phenomenon is close to a permeation flow).

\begin{figure*}[t]
\centering
\includegraphics[width=0.3\textwidth]{disc_bp2_tk_scan_25k.png}
\includegraphics[width=0.3\textwidth]{disc_bp2_tk_scan_75k.png}
\includegraphics[width=0.3\textwidth]{disc_bp2_tk_scan_125k.png}
\caption{BPII in shear flow for different thermodynamic parameters. The
snapshots show the evolution of the network with the flow direction into/out of
the page.  The reference state (blue, $\tau=-0.5, \kappa=2$), a high
temperature state (red, $\tau=0.5, \kappa=2$) and a low chirality state
(green, $\tau=-0.5, \kappa=1$) are shown; BPII is metastable in the later
two parameter sets.
The strain rate is $\dot{\gamma}=7.8125\ex{-5}$ in all three cases.
The pictures show a central cubic section consisting of 8 unit cells cut
from a larger simulation with $n_x=4$.}
\label{fig2}
\end{figure*}
